---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The variables included in this dataset are:

* steps: Number of steps taken in a 5-minute interval (missing values = NA)
* date: The date when the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

```{r load_libraries, message=FALSE, warning=FALSE, echo=TRUE}
library(dplyr)
library(lattice)
```
Load data from source file. The source file is assumed to be stored and read from a specified local folder. 

```{r, echo=TRUE, results='markup'}
act <- read.csv("activity.csv")
summary(act)
act$date<- as.Date(act$date)
```

## What is mean total number of steps taken per day?
Calculate the total number of steps taken per day.
```{r, echo=TRUE, results='markup'}
SumSteps <- aggregate(steps~date, act, FUN = sum,na.rm=TRUE)
colnames (SumSteps) <- c("date", "steps")
head(SumSteps)
```

Plot a histogram of SumSteps.
```{r}
hist(x = SumSteps$steps,
     main = "Histogram of Total Daily Steps", 
     xlab = "Daily Steps", ylab = "Frequency",
     ylim = c(0,20),
     breaks= 12.5)
```

Calculate the mean and median of the total number of steps taken per day.
```{r, echo=TRUE, results='markup'}
#mean
mean(SumSteps$steps) 
#median
median(SumSteps$steps) 
```

## What is the average daily activity pattern?
Calculate average steps per interval.
```{r, echo=TRUE, results='markup'}
AvgSteps <- aggregate(steps~interval,act, FUN = mean ,na.rm=TRUE)
names(AvgSteps)[2] <- "mean_steps"
str(AvgSteps)
```

Plot a time series of AvgSteps.
```{r}
plot(x = AvgSteps$interval,
     y = AvgSteps$mean_steps,
     type = "l",
     main = "Time Series of Average Number of Steps Taken Daily,\n in 5-Minute Interval",      
     xlab= "Interval in 5-Minute Interval", ylab ="Average Steps Daily")
```

Report Interval with highest average.
```{r, echo=TRUE, results='markup'}
#highest average for an interval
AvgSteps[which.max(AvgSteps$mean_steps),]
```

## Imputing missing values
Calculate the total number of missing values.
```{r, echo=TRUE, results='markup'}
#sum of missing values
sum(is.na(act$steps))
``` 

Impute missing values with the respective mean values for each day.
```{r, echo=TRUE, results='markup'}
#positions of missing values
full_act <- act
nas <- which(is.na(act$steps))
#mean for each interval
mean_interval <- tapply(full_act$steps, full_act$interval, mean, na.rm=TRUE, simplify=TRUE)
#NAs replaced by mean for each interval
full_act$steps[nas] <- mean_interval[as.character(full_act$interval[nas])]
head(full_act)
```

Plot a histogram of total steps per day with the new imputed data set.
```{r, results='markup'}
#sum of steps per day
SumAct <-aggregate(steps~date, act, FUN=sum)
names(SumAct)[2] <- "sum_steps"
str(SumAct)
#sum of missing values
sum(is.na(SumAct$sum_steps))
hist(SumAct$sum_steps,  
     main="Histogram of Total Daily Steps (Imputed)",
     xlab="Total Daily Steps",
     ylab="Frequency",
     ylim=c(0,25), breaks = 10)
```

Calculate the mean and median of the new imputed data set.
```{r, echo=TRUE, results='markup'}
#mean
mean(SumAct$sum_steps)  
#median
median(SumAct$sum_steps) 
```

* By imputing the missing data with the average number of steps in the same 5-min interval, both the mean and the median are equal to the same value: 10766.19. This is because the inserted average number brings the median closer to the mean. 

* Compared to the first part of the assignment, there is a slight change to the median (1 step difference).

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable with 2 levels; weekend and weekday.
```{r, echo=TRUE, results='markup'}
full_act <- mutate(full_act, daytype = ifelse(weekdays(full_act$date) == "Saturday" | weekdays(full_act$date) == "Sunday", "weekend", "weekday"))
full_act$daytype <- as.factor(full_act$daytype)
str(full_act)
```

Plot a panel plot of the 5 minute interval and the average number of steps taken, averaged across all weekday or weekend days.
```{r, echo=TRUE, results='markup'}
NewAvg <- aggregate(full_act$steps,list(interval=full_act$interval,weekdays=full_act$daytype),mean)
names(NewAvg) <- c("interval", "daytype", "mean")

xyplot(mean ~ interval | daytype, NewAvg, 
       type="l", 
       layout=c(1,2),
       main="Time Series of Average Number of Steps Taken Daily,\n Averaged Across All Weekend or Weekdays",
       xlab="5 Minute Interval", 
       ylab="Average Number of Steps"
)
```
